# Complement scoring {#complement-scoring}

```{r setup-9, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
```

The complement scoring was obtained from the file `/extdata-github/FINAL.Revised.BilatStudy.Path.MACstaining data.FINAL.xlsx`. The scoring system quantified the MAC staining into four levels:

1. No MAC staining of capillaries
2. Slight MAC staining
3. Moderate MAC staining
4. Prominent MAC staining 

However, no bilateral sample was classified as level 4.

```{r load-datasets, message=FALSE}
library(tidyverse)
library(corrr)
pkg_dir <- "/Users/cwon2/CompBio/Wellstone_BiLateral_Biopsy"
load(file.path(pkg_dir, "data", "comprehensive_df.rda"))
```

## Complement scoring (MCA score) and basket scores
First, we investigated the relationship between the complement scoring and the basket scores.

### Correlation with basekt scores

The complement scoring has moderate correlation with basket scores. The code chunk below calculates the Spearman correlation coefficients.

```{r get-complement-and-basket-score, message=FALSE}
# get complement scoring and basekt scores from  `comprehensive_df`
df <- comprehensive_df %>% 
  drop_na(`Complement Scoring`, `IG-logSum`) %>%
  dplyr::select(`Complement Scoring`, `DUX4-M6-logSum`,
                `IG-logSum`, `Complement-logSum`,
                `Inflamm-logSum`, `ECM-logSum`) 
  
  #dplyr::mutate(`Complement Scoring` = factor(`Complement Scoring`)) 
```
```{r complement-correlation-with-basket-score, message=FALSE}
df %>%
  corrr::correlate(., method="spearman") %>%
  dplyr::select(term, `Complement Scoring`) %>%
  dplyr::filter(term != "Complement Scoring") %>%
  dplyr::rename(basket = term) %>%
  dplyr::mutate(basket = str_replace(basket, "-logSum", "")) 
```
### Visualize the relationship 
```{r tidy-up-df}
# tidy up the data.frame for visualization
tidy_df <- df %>%
  dplyr::mutate(`Complement Scoring` = factor(`Complement Scoring`)) %>%
  tidyr::gather(key=`basket`, value=`basket score`, 
                -`Complement Scoring`) %>%
  dplyr::mutate(basket = str_replace(basket, "-logSum", "")) %>%
  dplyr::mutate(`Complement Scoring` = factor(`Complement Scoring`),
                basket = factor(basket, 
                                levels=c("DUX4-M6", "ECM", "Inflamm",
                                         "Complement", "IG"))) 
```

Display dotplot with color:
```{r dotplot-with-color, message=FALSE}
ggplot(tidy_df, aes(x=`Complement Scoring`, y=`basket score`)) +
    geom_dotplot(aes(fill=`Complement Scoring`,
                     color=`Complement Scoring`), 
                 binaxis='y', stackdir='center',
                 show.legend = FALSE,
                 stackratio=1.5, dotsize=1.5) +
    stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
                 geom="pointrange", color="red", fill="red", shape=23,
                 size=0.5, alpha=0.5) +
    facet_wrap(~basket, scales="free_y") +
    scale_fill_brewer(palette="Blues") +
    scale_color_brewer(palette="Blues") +
    theme_classic() + labs(x="complement scoring")
```

Make dotplot with grey scale for manuscript:
```{r dotplot-with-grey-scale, eval=FALSE, echo=FALSE}
ggplot(tidy_df, aes(x=`Complement Scoring`, y=`basket score`)) +
    geom_dotplot(aes(fill=`Complement Scoring`,
                     color=`Complement Scoring`), 
                 binaxis='y', stackdir='center',
                 show.legend = FALSE,
                 stackratio=1.5, dotsize=1.5) +
    stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
                 geom="pointrange", color="grey15", fill="grey15", shape=23,
                 size=0.5) +
    facet_wrap(~basket, scales="free_y") +
    scale_fill_manual(values=c("#cccccc", "#969696", "#636363", "#252525")) +
    scale_color_manual(values=c("#cccccc", "#969696", "#636363", "#252525")) +
    theme_classic() + labs(x="complement scoring")
# color for 4 levels: values=c("#cccccc", "#969696", "#636363", "#252525")
ggsave(file.path(pkg_dir, "manuscript", "figures",
                 "ComplementScoring-vs-basket-scores.pdf"))

```
ANOVA p-values for the null hypothesis that there is no difference between the main values of basket score in three levels of complement scoring.

```{r anova-complement-scoring-vs-basket-scores}
tm <- tidy_df %>% group_by(basket) %>% 
  dplyr::do(aov = summary(aov(`basket score` ~ `Complement Scoring`, 
                             data = .)))
names(tm$aov) <- levels(tidy_df$basket)
tm$aov
```


## Complement Scoring vs. STIR status
The average complement scoring for STIR- and STIR+ samples are 1.38 (+/-0.63SD) and 2.08 (+/-0.929SD), respectively.

```{r complement-scoring-vs-STIR, fig.align='center'}
comprehensive_df %>% 
  drop_na(`Complement Scoring`, `STIR_status`) %>%
  group_by(STIR_status) %>%
  summarize(mean=mean(`Complement Scoring`),
            sd = sd(`Complement Scoring`)) %>%
  ggplot(aes(x=STIR_status, y=`mean`)) + 
  geom_bar(stat="identity",fill="skyblue", alpha=0.7, 
           position=position_dodge()) +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.4,
                 position=position_dodge(.9), 
                colour="orange", alpha=0.9, linewidth=1.5) +
  theme_minimal() +
  theme(panel.grid.major = element_blank()) +
  labs(y="Complement Scoring", x="")
ggsave(file.path(pkg_dir, "manuscript", "figures",
                 "ComplementScoring-vs-STIR-status.pdf"),
       width=2, height=3)  
```

## Complement scoring vs. class
The average complement scoring for Control-like, Moderate+, and Muscle-Low samples are 1.15 (+- 0.376SD), 1.8 (+/- 0.894SD), and 1, respectively.

```{r complement-scoring-vs-class, fig.align='center'}
comprehensive_df %>% 
  drop_na(`Complement Scoring`, `class`) %>%
  dplyr::select(`Complement Scoring`, `class`) %>%
  group_by(class) %>%
  summarize(mean=mean(`Complement Scoring`),
            sd = sd(`Complement Scoring`)) %>%
  ggplot(aes(x=class, y=`mean`)) + 
  geom_bar(stat="identity",fill="skyblue", alpha=0.7, 
           position=position_dodge()) +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.4,
                 position=position_dodge(.9), 
                colour="orange", alpha=0.9, linewidth=1.5) +
  theme_minimal() +
  theme(panel.grid.major = element_blank()) +
  labs(y="Complement Scoring", x="")
ggsave(file.path(pkg_dir, "figures", "ComplementScoring-vs-FSHD-class.pdf"),
       width=3, height=3)
```