# (APPENDIX) Appendix {-}
# Build Ensembl-based annotation packages {#appendixA-buildTxDB}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Build _EnsDb_ annotation packages from `AnnotationHub`
Code below demonstrates a quick way for creating an _EnsDb_ annotation package using  the `AnnotationHub` package and the `GenomicFeatures::makeEnsembldbPackage()` function. For comprehensive information on Ensembl-based annotation packages, please refer to [Ensembldb](https://bioconductor.org/packages/release/bioc/html/ensembldb.html) on Bioconductor. 

```{r annotaitonHub-example, eval=FALSE}
#'
#' EnsDb.Hsapiens.v92: 
#'
library(AnnotationHub)
library(ensembldb)
library(GenomicFeatures)
ah <- AnnotationHub(localHub=TRUE)
query(ah, c("hsapiens"))
edb <- ah[["AH60977"]] # select the source GRF from ensembl version 92
seqlevelsStyle(edb) <- "NCBI"
GenomicFeatures::makeEnsembldbPackage(
  ensdb=dbfile(dbconn(edb)),
  version="1.0.0",
  maintainer="Chao-Jen Wong <cwon2@fredhutch.org>",
  author="Chao-Jen Wong",
  destDir="/fh/fast/tapscott_s/CompBio/hg38")
```


